Human capital mediates natural selection in contemporary humans

### 2022-03-14 why-natural-selection.Rmd
Response to BG reviewers: - add response document - move to RLRS not absolute N children   - some changes to over-time results - add analysis for respondents by Townsend deprivation - add bootstrap for effect of natural selection on income-EA3 R2, in FE model - add graph of mediation analysis - tweak plot style - other tweaks

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

@@

To test more directly for natural selection, we regress respondents' number of
children ($N_i$) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] Note that equation
\eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
score weighted by the number of children, which is the average polygenic score 
in the next generation.

@@

We also check for balancing and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
we find diversifying selection for educational attainment polygenic scores: at
higher values of these scores, the negative effect on fertility is smaller
(Appendix Figure \@ref(fig:plot-purifying)).

[^balance-diverse-def]: Balancing selection selects for intermediate values, 
reducing the variance of the trait under selection. Diversifying selection
selects for extreme values and increases the variance.

@@

```{r calc-siblings-children}
sib_chn_cor <- cor.test(famhist$n_sibs, famhist$n_children, use = "complete")
stopifnot(sib_chn_cor$p.value < 2e-16)
```

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings. Effect sizes of polygenic scores are highly
correlated across the two generations. Median-splitting respondents by year of
birth, we find little evidence of change in effect sizes among either the
parents' or the respondents' generation. We also check whether selection effects
vary by socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between *number of
siblings* and *number of children* is positive ($\rho$ = 
`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with our corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

effect_size <-  res_all %>% 
                  filter(dep.var == "N children", reg.type == "controlled") %>% 
                  pull(estimate)

@@

Second, we run a mediation analysis to directly test whether the correlation
between each polygenic score and fertility is mediated by educational attainment
(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$). Among these scores, the
median proportion of the total effect explained by the indirect effect is 
`r percent(median_prop)`. The educational attainment variable is a relatively
crude measure of human capital: more accurate measures would likely
explain more of the total effect.

@@

res_all %<>% filter(dep.var == "N children", reg.type == "controlled")
res_fe_fertility %<>% filter(Regression == "Raw")
res_all %<>% left_join(res_fe_fertility, by = "term")

@@

res_compare <- res_all %>% 
      filter(dep.var == "N children", reg.type == "controlled") %>% 
      left_join(res_risk_control %>% filter(term == score_name), 
                  by = "score_name")

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, most effect sizes appear unchanged over a large
span of twentieth-century history (Appendix Tables
\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. (Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital.) A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our risk-aversion parameter $\sigma$, which affects fertility by changing the
marginal utility of income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood,
and more lifetime sexual partners. Among mothers aged 22+, selection effects
are reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for these findings.
Because of the substitution effect of earnings on fertility, scores are selected
for when they correlate with low human capital, and this effect is stronger at 
lower levels of income and education.

@@

\FloatBarrier
\clearpage
## Balancing and diversifying selection

Balancing selection reduces variance in the trait under selection, while
diversifying selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
balancing selection ($p < 0.05/33$, negative coefficient on quadratic term).
Scores for educational attainment (EA2 and EA3) show significant diversifying
selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
selection against educational attainment at very high levels of the PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for balancing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings.
Scores for EA2 and EA3 again show significant diversifying selection ($p <
0.05/33$, positive coefficient). Other scores including hip circumference were 
not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 3}

@@

stopifnot(sig == c("EA2_noUKB", "EA3_excl_23andMe_UK", "hip_combined"))

@@

op <- par(mfrow = c(1, 3))
for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted N children"
       )
}
par(op)

@@

The UK Biobank data contains information on respondents' number of
siblings, i.e. their parents' number of children. Since respondents' polygenic 
scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has an additional source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For a clean comparison with the respondents' generation, we
rerun regressions on *number of children* excluding those with no children, and
show results in the figure. Selection effects are highly correlated across the
two generations, and most share the same sign. Absolute effect size estimates
are larger for the parents' generation. We treat this result cautiously, because
when we split respondents up by year of birth, we find few differences in effect
sizes between early- and late-born respondents, for either generation. In other
words, since estimated effect sizes change across "generations", but do not
change over time within either generation, the change may be due to remaining
ascertainment bias within the sample or other effects. In particular, effect
sizes in both generations may depend on polygenic scores' correlation with
childlessness, and we cannot estimate this for the parents' generation.

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
However, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`.

@@

stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 33)
stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 4)
stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 28)
```

@@

As an alternative approach, we run regressions interacting polygenic scores with
birth year, median split at 1950 ("early born" versus "late born"). We use both
*number of children* and *number of siblings* as a dependent variable. We weight
using age/qualification cells, and further adjust for selection in the parents'
generation (see above).

Tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period) summarize the results. There is no
evidence for changes in selection effects within the parents' generation. In the
respondents' generation, effect sizes were significantly larger in absolute size 
among the later-born for four polygenic scores: cognitive ability, EA2, EA3 and
extraversion. These changes are inconsistent with the intergenerational change, 
where estimated effect sizes were larger among the earlier, parents' generation.
One score, conscientiousness, showed a significant change in sign, from 
negative to positive effects on fertility.

Overall, there is weak evidence for change over time. The clearest results are
that (a) the direction of selection, and (b) the pattern of relative effect sizes
across scores, are broadly consistent over time.

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings). Weights: Age/Qualification times 1/number of siblings.")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children). Weights: Age/Qualification.")
```

@@

Figure \@ref(fig:plot-siblings-townsend) plots effects on *number of siblings* by
Townsend deprivation quintile of birth area.

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area. Higher = more deprived. Estimates weighted by 1/number of siblings.", fig.align = "center"}

@@

@@

res_sibs_unc <- filter(res_all, dep.var == "N siblings", reg.type == "controlled")
compare_abp_unc <- res_age_birth_parents %>% 
      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
      left_join(res_sibs_unc, by = "term")

@@

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on *number of siblings* on these
subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
by terciles of age at first live birth, for mothers and fathers. As in the
respondents' generation, effect sizes are smaller, or even oppositely signed, for older parents. Importantly, this holds for both sexes.

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
RLRS_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

RLRS is defined as respondents' number of children, divided by the mean number
of children of people born in the same year. The "selection effect", $\beta$,
reflects the strength of natural selection within the sample. In fact, since
polygenic scores are normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].[^normalization] Note that
equation \eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

[^normalization]: XXX The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
score weighted by the number of children, which is the average polygenic score 
in the next generation [@robertson1966mathematical].

@@

We also check for stabilizing and disruptive selection by estimating
\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
we find disruptive selection for educational attainment polygenic scores: at
higher values of these scores, the negative effect on fertility is smaller
(Appendix Figure \@ref(fig:plot-purifying)).

[^balance-diverse-def]: Stabilizing selection selects for intermediate values, 
reducing the variance of the trait under selection. Disruptive selection
selects for extreme values and increases the variance.

@@

```{r calc-siblings-children}
sib_chn_cor <- cor.test(famhist$RLRS_parents, famhist$RLRS, use = "complete")
stopifnot(sib_chn_cor$p.value < 2e-16)
```

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among either the parents' or
the respondents' generation. We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1, and let $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
of the utility function, i.e. the decline in marginal utility of income as income
increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with our corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of $\sigma$, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

effect_size <-  res_all %>% 
                  filter(dep.var == "RLRS", reg.type == "controlled") %>% 
                  pull(estimate)

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$) and takes the same sign as the
overall effect. Among these scores, the median proportion of the total effect
explained by the indirect effect is `r percent(median_prop)`. Figure
\@ref(fig:plot-mediation) shows estimated proportions explained along
with bootstrap 95% confidence intervals (uncorrected; 100 bootstraps).
The educational attainment variable is a relatively crude measure of human 
capital: more accurate measures would likely explain more of the total effect.

```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, polygenic scores with significant selection effects"}

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/33` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey30"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

res_all %<>% filter(dep.var == "RLRS", reg.type == "controlled")
res_fe_fertility %<>% filter(Regression == "Raw")
res_all %<>% left_join(res_fe_fertility, by = "term")

@@

res_compare <- res_all %>% 
      filter(dep.var == "RLRS", reg.type == "controlled") %>% 
      left_join(res_risk_control %>% filter(term == score_name), 
                  by = "score_name")

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, most effect sizes appear unchanged over a large
span of twentieth-century history (Appendix Tables
\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood,
and more lifetime sexual partners. Among mothers aged 22+, selection effects
are reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. Strikingly,
some of these results were predicted by @fisher1930genetical, pp. 253-254.

The economic theory of fertility provides a parsimonious explanation for these
findings. Because of the substitution effect of earnings on fertility, scores
are selected for when they correlate with low human capital, and this effect is
stronger at lower levels of income and education.

@@

\FloatBarrier
\clearpage
## Stabillizing and disruptive selection

Stabilizing selection reduces variance in the trait under selection, while
disruptive selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
The EA2 score for educational attainment shows significant disruptive
selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
selection against educational attainment at very high levels of the PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for stabilizing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings.
Scores for EA2 and EA3 show significant disruptive selection ($p <
0.05/33$, positive coefficient on quadratic). Other scores including 
hip circumference were not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score.", fig.height = 3}

@@

stopifnot(sig == c("EA2_noUKB", "hip_combined"))

@@

op <- par(mfrow = c(1, 2))
for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted RLRS"
       )
}
par(op)

@@

The UK Biobank data contains information on respondents' number of siblings
(including them), i.e. their parents' number of children. Since respondents'
polygenic scores are equal in expectation to the mean scores of their parents,
we can use this to look at selection effects in the parents' generation. We
estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
variable. The parents' generation has an additional source of ascertainment
bias: sampling parents of respondents overweights parents who have many
children. For instance, parents of three children will have, on average, three
times more children represented in UK Biobank than parents of one child. Parents
of no children will by definition not be represented. To compensate, we reweight
our preferred weightings (Age/Qualification) by the inverse of *number of
siblings*.

Figure \@ref(fig:plot-res-sibs) shows regressions of parents' RLRS on
polygenic scores. For a clean comparison with the respondents' generation, we
rerun regressions on respondents' RLRS excluding those with no children, and
show results in the figure. Selection effects are highly correlated across the
two generations, and most share the same sign. Absolute effect size estimates
are larger for the parents' generation. We treat this result cautiously, because
effect sizes in both generations may depend on polygenic scores' correlation with
childlessness, and we cannot estimate this for the parents' generation.

To learn more about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
However, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`.

@@

stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 32)
stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 9)
stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 23)
```

@@

As an alternative approach, we run regressions interacting polygenic scores with
birth year, median split at 1950 ("early born" versus "late born"). We use both
respondents' RLRS and parents' RLRS as a dependent variable. We weight
using age/qualification cells, and further adjust for selection in the parents'
generation (see above).

Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
summarize the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
ability, coronary artery disease, EA2, EA3, extraversion and major depressive
disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.
One score, for alzheimer's disease, showed a significant change in sign, from
negative to positive effects on fertility.

Overall, there is some evidence for an increase in the strength of selection in
recent history. The clearest result is that the pattern of relative effect sizes
across scores is broadly consistent over time.

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on parents' RLRS). Weights: Age/Qualification times 1/number of siblings.")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on RLRS). Weights: Age/Qualification.")
```

@@

Figure \@ref(fig:plot-siblings-townsend) plots effects on parents' RLRS by
Townsend deprivation quintile of birth area.

```{r plot-siblings-townsend, fig.cap = "Selection effects (parents' RLRS) by Townsend deprivation quintile of birth area. Higher = more deprived. Estimates weighted by 1/number of siblings.", fig.align = "center"}

@@

For comparison, Figure \@ref(fig:plot-townsend) plots effects on 
respondents' RLRS by Townsend deprivation quintile of birth area.

```{r plot-townsend, fig.cap = "Selection effects in the respondents' generation by Townsend deprivation quintile of birth area. Higher = more deprived.", fig.align = "center"}

drake::loadd(res_townsend)
n_regs <- nrow(res_townsend)

res_townsend %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5
                  )

@@

res_sibs_unc <- filter(res_all, dep.var == "RLRS_parents", reg.type == "controlled")
compare_abp_unc <- res_age_birth_parents %>% 
      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
      left_join(res_sibs_unc, by = "term")

@@

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on parents' RLRS on these
subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
by terciles of age at first live birth, for mothers and fathers. As in the
respondents' generation, effect sizes are smaller, or even oppositely signed, for older parents. Importantly, this holds for both sexes.

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) by age at first live birth terciles.  Estimates weighted by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
